---
title: "Native Plant Color Palette"
author: "Kassandra Jensen"
date: "2024-12-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Libraries
```{r}
library(ggplot2)
```

## Plant colors
I am making a package based on the native plants found throughout my surveys. The colors were extracted from plant photos using [this website](https://redketchup.io/color-picker#google_vignette). Each of the plants will have 5 associated colors. 
```{r}
#Make a vector for each plant:

#Cyanea hirtella
CYAHIR <- c(
  "inner_flower" = "#CB81B2",
  "outer_flower" = "#F8FAF9",
  "stigma" = "#C9C468",
  "leaf" = "#A7CB0F",
  "leaf2" = "#565A10"
)

#Touchardia latifolia
TOULAT <- c(
  "berry" = "#E67637",
  "fruit" = "#404E2A",
  "leaf" = "#638D27",
  "stem" = "#A2A94C",
  "vein" = "#7B3926"
)

#Hibiscus waimeae subsp. hannerae
HIBWAIHAN <- c(
  "petals" = "#FDFDFD",
  "stigma" = "#FD3C5B",
  "pollen" = "#D5953E",
  "leaf_rust"= "#C95646",
  "leaf" = "#77A337"
)

#Pipturus kauiensis
PIPKAU <- c(
  "leaf" = "#80AD46",
  "vein" = "#A74A55",
  "fruit" = "#DEE0DD",
  "seed" = "#1C201F",
  "underside_leaf" = "#FDFFF2"
)

#Kadua acuminata
KADACU <-c(
  "leaf" = "#478A21",
  "leaf2" = "#96879E",
  "leaf3" = "#85C100",
  "flower" = "#D2EAC8",
  "anther" = "#1D1307"
)

#Boehmeria grandis
BOEGRA <- c(
  "leaf" = "#779D82",
  "inflorescence" = "#B62552",
  "flower" = "#DADEE8",
  "vein" = "#9B4155",
  "underside_leaf" = "#85A04D"
)

#Psychotria mariniana
PSYMAR <- c(
  "leaf" = "#72A03D",
  "fruit" = "#F99F2F",
  "unripe_fruit" = "#73A301",
  "flower" = "#E7E7E7",
  "bark" = "#483A37"
)

#Vitex rotundifolia
VITROT <- c(
  "flower" = "#9A71F3",
  "leaf" = "#C8DEBA",
  "filament" = "#A893D4",
  "hairs" = "#FAFFFB",
  "flower2" = "#B98FF1"
)

#Cyrtandra sp.
CYRTANDRA <- c(
  "leaf" = "#91C44E",
  "leaf2" = "#74A35D",
  "flower" = "#CFE1D1",
  "flowerparts" = "#41361A",
  "seed" ="#DDAD59"
)

```

# Make a list of all the palette names
```{r}
plant_palettes <- list(
  CYAHIR = CYAHIR,
  TOULAT = TOULAT,
  HIBWAIHAN = HIBWAIHAN,
  PIPKAU = PIPKAU,
  KADACU = KADACU,
  BOEGRA = BOEGRA,
  PSYMAR = PSYMAR,
  VITROT = VITROT,
  CYRTANDRA = CYRTANDRA
)
```

# Function
This is a function so that you can use the six letter species code to find the associated color pallete. 
```{r}
plantpalette <- function(palette_name) {                 ## Make function called plantpalette, manipulating "palette_name" variable
  if (!palette_name %in% names(plant_palettes)) {        ## calls plant_palettes to look for name, %in% checks if the palette name exists, ! makes it so if not found, the "!" makes it a negative statement so that if the plant name isn't found in the list, then the stop message is given
    stop("Palette not found. Available palettes: ",     ## stop: stops function to give the error, paste: puts the names as a sting but separates by a comma
         paste(names(plant_palettes), collapse = ", "))
  }
  return(plant_palettes[[palette_name]])                ## accesses the palette with the name given and returns the palette!
}

## Example
plantpalette("CYRTANDRA")                                  ##it worked! yay!
```
# Showcase the colors!
Made another function to quickly showcase the color palettes. 
```{r}
example_plot <- function(palette_name) {                    ## function name = example_plot, manipulating "palette_name" variable
  colors <- plantpalette(palette_name)                      ## gets the colors of the palette you name, uses the previous function
  par(bg = "lightgray")                                     ## this just makes the background grey so it's easier to see the lighter colors
  barplot(rep(1, length(colors)),                           ## makes a barplot, with a vector of 1s with the amount of colors in the package
          col = colors,                                     ## calls the colors in the package called
          border = NA,                                      ## remove borders so it looks nicer
          main = palette_name)                              ## makes the title the plant code
}

## Test all palettes using a loop! Thankfully we just learned this! lol
## Calls a palette then loops it through the function
for (palette in names(plant_palettes)) {                    ## calls all the names in plant_palettes, loops it for each palette 
  example_plot(palette)                                     ## calls the function above
}

```